A Standard for the Scholarly Citation of Archaeological Data

### 2017-09-17 analysis/paper/paper.Rmd
second submit

---
title: "A Standard for the Scholarly Citation of Archaeological Data"
author:
  - "Ben Marwick, Department of Anthropology, University of Washington, Seatll, 98115, WA, USA, bmarwick@uw.edu, corresponding author"
  - "Suzanne E. Pilaar Birch, Anthropology and Geography Department, University of Georgia,  Athens, Georgia,  30602, USA, sepbirch@uga.edu"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
    fig_caption: yes
    reference_docx: ../templates/template.docx
abstract: |
  "How do archaeologists share their research data, if at all? We review what data are, according to current influential definitions, and previous work on the benefits, costs and norms of data sharing in the sciences broadly. To understand data sharing in archaeology, we present the results of three pilot studies: requests for data by email; review of data availability in published articles, and analysis of archaeological datasets deposited in repositories. We find that archaeologists are often willing to share, but discipline-wide sharing is patchy and ad hoc. Legislation and mandates are effective at increasing data-sharing, but editorial policies at journals lack adequate enforcement. Although most of data available at repositories are licensed to enable flexible reuse, only a small proportion of the data are stored in structured formats for easy reuse. We present some suggestions for improving the state of date sharing in archaeology, among these is a standard for citing data sets to ensure that researchers making their data publicly available receive appropriate credit."
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

# Introduction

@@

Pragmatically in our own work, we consider that the main data suitable for sharing as the contents of the computer files (e.g. text, raw structured data, and images) that we analyse or otherwise use to create the figures, tables and quantitative results that we include in our reports and publications. The selection omits materials that could also be considered data accoring to the defintions noted above (e.g. field notes), but we feel our selection represents the most information-dense and meaningufl data records that are useful for others wanting to assess the reliability of our work, or resuse our findings in their own work. Our citation standard imposes no constraints on what data can be, but we propose a few minimal qualities for how data should be organised that will help ground our discussion. Table \@ref(tab:desirable) summarises some desirable minimum qualities for archaeological data that optimise its discovery and reuse.

@@

The surveys above show that data sharing practices tend to be a private activity, but where public sharing does occur, what forms does it take, and how do these vary across different fields of research?  Wouters and Haak (2017) surveyed 1,200 researchers about data sharing behaviours and found a spectrum from ‘intensive data-sharing fields’ to ‘restricted data-sharing fields’. Intensive data-sharing fields often depend on access to global databases, for example human genetics, while restricted data-sharing fields depend mostly on data collected by the individual researcher or their immediate collaborators, such as digital humanities. These differences in disciplinary norms have implications for data ownership (intensive: collectively owned, willing to share; restrictive: owned by the individual, reluctant to share) and the effort required to share (intensive: low effort because data are prepared for sharing as an integral part of the research process, restrictive: high effort because data are prepared for sharing after the research, or apart from it, in response to a request from another researcher). This intensive-restricted spectrum also maps on to Sawyer's (2008) analysis of data sharing in data-rich fields compared to data-poor fields. Data rich fields, such as astronomy, oceanography, and ecology, are have three common characterics: pooling and sharing data resources is expected; the cumulative nature of work that builds on common data and common methods reduces the likelihood of alternative data sources and other methods being common; and the number of theoretical choices is limited because there is enough data to adequately test competing theories. Data-poor fields have three common characteristics: data are rare, or hard to get, or both; the types of data available often dictate the methods taken; and theories are relatively easy to generate and more difficult to validate. In data-poor fields, researchers often invest substantial time and resources on gathering their own data. The getting of data, and the quality of the data collected, is often a key determinant of the status of a researcher, leading to hoarding of data as an asset to be traded for professional benefit.

@@

A DOI alone is not sufficient for citing a dataset because, like many things onthe internet, the files can change over time. For example we might upload data files to the repository when we submit a paper for peer review, then get a DOI and cite that in our manuscript so the peer reviewers can see them (some services provide a private URL that can be shared with reviewers but is not publicly accessible or discoverable by a search engine). Then when we resubmit our manuscript after peer review, we might update the repository to include additional data files relating to additional analyses requested by the reviewers. We still have the same DOI in our manuscript for this data set, but the files have changed because of the revisions we did in response to the peer reviewers' comments. Most data repositories will keep each version of the files that you upload, so in this scenario there will be at least two versions of the files in our data repository. Even if we do not follow this exact sequence of updating the files, we may still change them in the future, after our paper has been published. Because of this potential to change the files stored at the DOI, it is necessary to identify specific file versions in our citation of the data. A simple method to record the version is to include the date that the data files were accessed, as when citing any other resource found on the internet. Some repositories also provide DOIs that have a version number suffix like '.v2', '.2', '/2'. However this is not formally a functionality of the DOI system, so including the date accessed is a more robust and general method of indicating the version. Technological solutions to issues of transience, permanence and granularity in data citation are still in development.

@@

In short, there is more to the ethics of data sharing than avoiding risk of damage. Redacting the locations of archaeological sites to prevent looting is an obvious case of preparing data for sharing to reduce the risk of damage to the sites (and in some juridictions, mandated by law). However, because of cultural, economic and political differences between Western archaeologists and the communities they work with, an archaeologist cannot consider their own risk assessment to be comprehensive and exhaustive (Nicholas and Wylie 2012). It is crucial for archaeologists to discuss data sharing plans during negotiations with representatives authorized by the indigenous peoples whose cultural heritage is the subject of investigation. This negotiation process is important because only indigenous peoples themselves can identify potential adverse outcomes to data sharing, and they can do this only if they understand the proposed research and anticipated results. The concrete outcome of this process is often a Memorandum of Understanding between the indigenous community and the archaeologist's employer that documents the outcome of negotiations about data sharing. If the archaeologist's indigenous partners have reasons not to share data in public, then the archaeologist should respect that position, and note in publication that the data are restricted at the request of the indigenous collaborators.

@@

Among the many challenges of increasing the rate of data sharing in archaeology and other 'small' sciences is that data sharing is unfunded, unrewarded and is only rarely required. The challenge of funding can be addressed by professional societies, such as the Society of American Archaeology, communicating to funding bodies, such as the National Science Foundation, about the importance of allocating funds specifically for data sharing and requesting a chance in policy. The challenge of requiring data sharing can also be addressed by funding bodies, but may be most effectively tackled by journal editors and editorial boards, for example by adopting the Transparency and Openness Protocols of the Centre for Open Science (Nosek et al. 2015). Changes in journal-specific submission guidelines can result in desirable changes in authors’ research practices (Giofrè et al 2017). For example, if archaeology journals adopt incentives such as badging (Kidwell et al. 2016), and require as a condition of publication that submitted articles include in the text a properly formatted citation to the data analysed in the article (or an explanation why the data are not available), authors would quickly adapt to this new requirement to ensure their article will be published. As peer reviewers, we can also communicate to editors that we require the manuscripts that we review to include citations of datasets analysied in the manuscript.

@@

This paper emerged out of discussions at the 2017 SAA meeting, specifically the forum ‘Beyond Data Management: A Conversation about “Digital Data Realities”’, organised by Sarah Kansa and Eric Kansa, and the SAA publications committee meeting. Thanks to Canan Cakirlar for insights into the history of archaeological data sharing in the Netherlands.

@@

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

>>>>

---
title: "A Standard for the Scholarly Citation of Archaeological Data"
author: 
  - name          : "Ben Marwick"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Department of Anthropology, University of Washington, Seattle, 98115, WA, USA"
    email         : "bmarwick@uw.edu"
  - name          : "Suzanne E. Pilaar Birch"
    affiliation   : "2"

affiliation:
  - id            : "1"
    institution   : "University of Washington, Seattle, Washington, USA"
  - id            : "2"
    institution   : "University of Georgia,  Athens, Georgia,  USA"

date: "`r format(Sys.time(), '%d %B, %Y')`"

figsintext        : no
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : no
keywords          : ""
wordcount         : ""

lang              : "english"
class             : "man"

output: 
  papaja::apa6_word

abstract: |
  "How do archaeologists share their research data, if at all? We review what data are, according to current influential definitions, and previous work on the benefits, costs and norms of data sharing in the sciences broadly. To understand data sharing in archaeology, we present the results of three pilot studies: requests for data by email; review of data availability in published articles, and analysis of archaeological datasets deposited in repositories. We find that archaeologists are often willing to share, but discipline-wide sharing is patchy and ad hoc. Legislation and mandates are effective at increasing data-sharing, but editorial policies at journals lack adequate enforcement. Although most of data available at repositories are licensed to enable flexible reuse, only a small proportion of the data are stored in structured formats for easy reuse. We present some suggestions for improving the state of date sharing in archaeology, among these is a standard for citing data sets to ensure that researchers making their data publicly available receive appropriate credit."
---

@@

@@

Pragmatically in our own work, we consider that the main data suitable for sharing as the contents of the computer files (e.g. text, raw structured data, and images) that we analyse or otherwise use to create the figures, tables and quantitative results that we include in our reports and publications. The selection omits materials that could also be considered data according to the definitions noted above (e.g. field notes), but we feel our selection represents the most information-dense and meaningful data records that are useful for others wanting to assess the reliability of our work, or reuse our findings in their own work. These are also the items that have been most frequently requested by others writing to us about our publications. Our citation standard imposes no constraints on what data can be, but we propose a few minimal qualities for how data should be organised that will help ground our discussion. Table \@ref(tab:desirable) summarises some desirable minimum qualities for archaeological data that optimise its discovery and reuse.

@@

The surveys above show that data sharing practices tend to be a private activity, but where public sharing does occur, what forms does it take, and how do these vary across different fields of research?  Wouters and Haak (2017) surveyed 1,200 researchers about data sharing behaviours and found a spectrum from ‘intensive data-sharing fields’ to ‘restricted data-sharing fields’. Intensive data-sharing fields often depend on access to global databases, for example human genetics, while restricted data-sharing fields depend mostly on data collected by the individual researcher or their immediate collaborators, such as digital humanities. These differences in disciplinary norms have implications for data ownership (intensive: collectively owned, willing to share; restrictive: owned by the individual, reluctant to share) and the effort required to share (intensive: low effort because data are prepared for sharing as an integral part of the research process, restrictive: high effort because data are prepared for sharing after the research, or apart from it, in response to a request from another researcher). This intensive-restricted spectrum also maps on to Sawyer's (2008) analysis of data sharing in data-rich fields compared to data-poor fields. Data rich fields, such as astronomy, oceanography, and ecology, are have three common characteristics: pooling and sharing data resources is expected; the cumulative nature of work that builds on common data and common methods reduces the likelihood of alternative data sources and other methods being common; and the number of theoretical choices is limited because there is enough data to adequately test competing theories. Data-poor fields have three common characteristics: data are rare, or hard to get, or both; the types of data available often dictate the methods taken; and theories are relatively easy to generate and more difficult to validate. In data-poor fields, researchers often invest substantial time and resources on gathering their own data. The getting of data, and the quality of the data collected, is often a key determinant of the status of a researcher, leading to hoarding of data as an asset to be traded for professional benefit.

@@

A DOI alone is not sufficient for citing a dataset because, like many things on the internet, the files can change over time. For example we might upload data files to the repository when we submit a paper for peer review, then get a DOI and cite that in our manuscript so the peer reviewers can see them (some services provide a private URL that can be shared with reviewers but is not publicly accessible or discoverable by a search engine). Then when we resubmit our manuscript after peer review, we might update the repository to include additional data files relating to additional analyses requested by the reviewers. We still have the same DOI in our manuscript for this data set, but the files have changed because of the revisions we did in response to the peer reviewers' comments. Most data repositories will keep each version of the files that you upload, so in this scenario there will be at least two versions of the files in our data repository. Even if we do not follow this exact sequence of updating the files, we may still change them in the future, after our paper has been published. Because of this potential to change the files stored at the DOI, it is necessary to identify specific file versions in our citation of the data. A simple method to record the version is to include the date that the data files were accessed, as when citing any other resource found on the internet. Some repositories also provide DOIs that have a version number suffix like '.v2', '.2', '/2'. However this is not formally a functionality of the DOI system, so including the date accessed is a more robust and general method of indicating the version. Technological solutions to issues of transience, permanence and granularity in data citation are still in development.

@@

In short, there is more to the ethics of data sharing than avoiding risk of damage. Redacting the locations of archaeological sites to prevent looting is an obvious case of preparing data for sharing to reduce the risk of damage to the sites (and in some jurisdictions, mandated by law). However, because of cultural, economic and political differences between Western archaeologists and the communities they work with, an archaeologist cannot consider their own risk assessment to be comprehensive and exhaustive (Nicholas and Wylie 2012). It is crucial for archaeologists to discuss data sharing plans during negotiations with representatives authorized by the indigenous peoples whose cultural heritage is the subject of investigation. This negotiation process is important because only indigenous peoples themselves can identify potential adverse outcomes to data sharing, and they can do this only if they understand the proposed research and anticipated results. The concrete outcome of this process is often a Memorandum of Understanding between the indigenous community and the archaeologist's employer that documents the outcome of negotiations about data sharing. If the archaeologist's indigenous partners have reasons not to share data in public, then the archaeologist should respect that position, and note in publication that the data are restricted at the request of the indigenous collaborators.

@@

Among the many challenges of increasing the rate of data sharing in archaeology and other 'small' sciences is that data sharing is unfunded, unrewarded and is only rarely required. The challenge of funding can be addressed by professional societies, such as the Society of American Archaeology, communicating to funding bodies, such as the National Science Foundation, about the importance of allocating funds specifically for data sharing and requesting a chance in policy. The challenge of requiring data sharing can also be addressed by funding bodies, but may be most effectively tackled by journal editors and editorial boards, for example by adopting the Transparency and Openness Protocols of the Centre for Open Science (Nosek et al. 2015). Changes in journal-specific submission guidelines can result in desirable changes in authors’ research practices (Giofrè et al 2017). For example, if archaeology journals adopt incentives such as badging (Kidwell et al. 2016), and require as a condition of publication that submitted articles include in the text a properly formatted citation to the data analysed in the article (or an explanation why the data are not available), authors would quickly adapt to this new requirement to ensure their article will be published. As peer reviewers, we can also communicate to editors that we require the manuscripts that we review to include citations of datasets analysed in the manuscript.

@@

This paper emerged out of discussions at the 2017 SAA meeting, specifically the forum ‘Beyond Data Management: A Conversation about “Digital Data Realities”’, organised by Sarah Kansa and Eric Kansa, and the SAA publications committee meeting. Thanks to Canan Cakirlar for insights into the history of archaeological data sharing in the Netherlands. BM was supported by an Australian Research Council Future Fellowship (FT140100101).

@@

This document contains `r prettyNum(wordcountaddin:::text_stats_fn_(readLines("paper.Rmd"))$n_words_korp, big.mark = ",")` words and was generated on `r Sys.time()` using the following computational environment and dependencies:
